Reasoning about a Capability Machine with Local Capabilities\\ Provably Safe Stack and Return Pointer Management (without OS Support)\\ Technical Appendix Including Proofs and Details

### 2017-12-18 presentations/illustrations/preamble.tex
preamble

\documentclass{beamer}

\usepackage{pgf}
\usepackage{ulem}
\usepackage{tikz}
\usetikzlibrary{backgrounds}
\usetikzlibrary{positioning}
\usetikzlibrary{patterns}
\usetikzlibrary{decorations.pathreplacing}
\usetikzlibrary{arrows,shapes}
\usepackage{tikzpeople}

@@

\begin{document}
\begin{frame}
  \frametitle{Traditional stack pointers}
  \begin{tikzpicture}[scale=.5, every node={scale=.5}]
    % recurrent parts
    \begin{scope}
      \clip (-.1,-.1) rectangle (6.1,15.1);
      \draw[fill=gray!20,draw=none,opacity=.8] (0,0) rectangle (6,15);
      \draw[draw=gray!80] (0,0) -- (0,15);
      \draw[draw=gray!80] (6,0) -- (6,15);
      \draw[fill=white] (0,-.5) rectangle (6,2.5) node[pos=.5] {\footnotesize higher stack frames...};
    \end{scope}
    \draw[->] (-1,0) -- node[midway,sloped,above] {stack grows upward} (-1,15);

    % traditional:
    \draw<+>[fill=white] (0,2.5) rectangle (6,5.5) node[pos=.5] {\footnotesize caller stack frame};
    \draw<.> (8,7) node[draw=black] (sp0) {};
    \draw<.> node[right=0cm of sp0] (sp0l) { stack pointer };
    \draw<.> (sp0) edge[->,out=235,in=0] (6,5.25);

    % traditional call:
    \draw<+>[fill=white] (0,2.5) rectangle (6,5.5) node[pos=.5] {\footnotesize caller stack frame};
    \draw<.>[fill=white] (0,5.5) rectangle (6,8.5) node[pos=.5] {\footnotesize callee stack frame};
    \draw<.> (8,7) node[draw=black] (sp0) {};
    \draw<.> node[right=0cm of sp0] (sp0l) { stack pointer };
    \draw<.> (sp0) edge[->,out=235,in=0] (6,8.25);

    % traditional return:
    \draw<+>[fill=white] (0,2.5) rectangle (6,5.5) node[pos=.5] {\footnotesize caller stack frame};
    \draw<.> (8,7) node[draw=black] (sp0) {};
    \draw<.> node[right=0cm of sp0] (sp0l) { stack pointer };
    \draw<.> (sp0) edge[->,out=235,in=0] (6,5.25);

    % traditional call, attack:
    \draw<+->[fill=white] (0,2.5) rectangle (6,5.5) node[pos=.5] {\footnotesize caller stack frame};
    \draw<.->[fill=white] (0,5.5) rectangle (6,8.5) node[pos=.5] {\footnotesize callee stack frame};
    \draw<.-> (8,7) node[draw=black] (sp0) {};
    \draw<.-> node[right=0cm of sp0] (sp0l) { stack pointer };

    % read other stack frames
    \draw<.> (sp0) edge[->,out=235,in=0] (6,8.25);
    \draw<.> (3,7) node[devil,mirrored,minimum size=1cm] {};
    \draw<.> (sp0) edge[->,color=red,very thick,out=235,in=0] node[sloped, below] {read/write} (6,4);

    % break well-bracketedness
    \draw<+>[fill=red,opacity=.7] (-.5,2.5) rectangle (6.5,8.5);
    \draw<.-> (8,7) node[draw=black] (sp0) {};
    \draw<.>[red] (sp0) edge[->,out=235,in=0] (6,1.75);
    \draw<.> node[color=red,align=left,below=of sp0l] {return but\\
      skip caller frame};
  \end{tikzpicture}

\end{frame}

\begin{frame}
  \frametitle{Stack and return capabilities}
  \begin{tikzpicture}[scale=.5, every node={scale=.5}]
    % recurrent parts
    \begin{scope}
      \clip (-.1,-.1) rectangle (6.1,15.1);
      \draw[fill=gray!20,draw=none,opacity=.8] (0,0) rectangle (6,15);
      \draw[draw=gray!80] (0,0) -- (0,15);
      \draw[draw=gray!80] (6,0) -- (6,15);
      \draw[fill=gray!50] (0,-.5) rectangle (6,2.5) node[pos=.5,color=black] {\footnotesize higher stack frames...};
    \end{scope}

    \draw[->] (-2,0) -- node[midway,sloped,above] {stack grows upward} (-2,15);

    % first step: capabilities
    \draw<+>[fill=white] (0,2.5) rectangle (6,5.5) node[pos=.5] {\footnotesize current stack frame};
    \begin{scope}
      \clip (-.1,-.1) rectangle (6.1,15);
      \draw<.>[fill=white] (0,5.5) rectangle (6,15.1);
    \end{scope}
    \draw<.> [decorate,decoration={brace,amplitude=10pt,mirror,raise=4pt},yshift=0pt]
    (6.5,2.5) -- (6.5,16) node[draw=black] (sp1) [black,midway,xshift=0.8cm] {};
    \draw<.> node[right=0cm of sp1] { stack pointer };
    \draw<.> [decorate,decoration={brace,amplitude=10pt,mirror,raise=4pt},yshift=0pt]
      (6,0) -- (6,16) node (oldframe1) [black,midway,xshift=0.5cm] {};
    \draw<.> node[fill=gray!50,right=0cm of oldframe1] (rp1l) {return pointer data};

    % call using capabilities
    \draw<+>[fill=gray!50] (0,2.5) rectangle (6,5.5) node[pos=.5] {\footnotesize caller stack frame};
    \draw<.>[fill=white] (0,6.5) rectangle (6,9.5) node[pos=.5] {\footnotesize callee stack frame};
    \begin{scope}
      \clip (-.1,-.1) rectangle (6.1,15);
      \draw<.>[fill=white] (0,9.5) rectangle (6,15.1);
    \end{scope}
    \draw<.> [decorate,decoration={brace,amplitude=10pt,mirror,raise=4pt},yshift=0pt]
      (6.5,6.5) -- (6.5,16) node[draw=black] (sp1) [black,midway,xshift=0.8cm] {};
    \draw<.> node[right=0cm of sp1] { stack pointer };
    \draw<.> [decorate,decoration={brace,amplitude=10pt,mirror,raise=4pt},yshift=0pt]
      (6,2.5) -- (6,16) node (callerframe1) [draw=black,midway,xshift=0.8cm] {};
    \draw<.> node[fill=gray!50,right=0cm of callerframe1] (rp1l) {return pointer data};
    \draw<.>[fill=gray!50] (0,5.5) rectangle node {\scriptsize old return pointer data} (6,6);
    \draw<.>[fill=gray!50] (0,6) rectangle node {\scriptsize old return pointer code} (6,6.5);
    \draw<.> [decorate,decoration={brace,amplitude=10pt,raise=4pt},yshift=0pt]
      (0,0) -- (0,16) node (oldframe1) [black,midway,xshift=-0.5cm] {};
    \draw<.> (0,5.75) edge[->,out=180,in=225] (oldframe1);

  \end{tikzpicture}
\end{frame}

\begin{frame}
  \frametitle{Stack and return capabilities: Attack 1}
  \begin{tikzpicture}[scale=.5, every node={scale=.5}]
    % recurrent parts
    \begin{scope}
      \clip (-.1,-.1) rectangle (6.1,15.1);
      \draw[fill=gray!20,draw=none,opacity=.8] (0,0) rectangle (6,15);
      \draw[draw=gray!80] (0,0) -- (0,15);
      \draw[draw=gray!80] (6,0) -- (6,15);
      \draw[fill=gray!50] (0,-.5) rectangle (6,2.5) node[pos=.5,color=black] {\footnotesize higher stack frames...};
      \begin{scope}
        \clip (-.1,-.1) rectangle (6.1,15);
        \draw[fill=white] (0,2.5) rectangle (6,15.1);
      \end{scope}
    \end{scope}

    \draw[->] (-2,0) -- node[midway,sloped,above] {stack grows upward} (-2,15);

    % attacker...
    \begin{onlyenv}<+-+(2)>
      \draw[fill=gray!50] (0,2.5) rectangle (6,6.5) node[pos=.5] {\footnotesize caller stack frame};
      \draw[fill=white] (0,6.5) rectangle (6,9.5) node[pos=.5] {\footnotesize callee stack frame};
      \draw (3,8) node[devil,mirrored,minimum size=1cm] {};
      \draw [decorate,decoration={brace,amplitude=10pt,mirror,raise=4pt},yshift=0pt]
      (6.5,6.5) -- (6.5,16) node[draw=black] (sp1) [black,midway,xshift=0.8cm] {};
      \draw node[right=0cm of sp1] { stack pointer };
    \end{onlyenv}

    % stores stack pointer in heap
    \begin{onlyenv}<+->
      \begin{scope}
        \clip (13.9,6) rectangle (18.1,13);
        \draw[fill=gray!20] (14,5.9) rectangle (18,13.1);
      \end{scope}
      \draw (16,14) node {heap memory};
      \draw[fill=gray!50] (14,8) rectangle node[color=red] {\scriptsize copy of old sp} (18,8.5);
    \end{onlyenv}
    \draw<.> (14,8.25) edge[->,red,thick,bend left] (sp1);
    \begin{onlyenv}<+->
      \draw [decorate,decoration={brace,amplitude=10pt,mirror,raise=4pt},yshift=0pt]
      (11.5,6.5) -- (11.5,16) node[draw=black] (spold) [black,midway,xshift=0.8cm] {};
      \draw (14,8.25) edge[->,red,thick,bend left] (spold);
    \end{onlyenv}

    % attacker returns
    \begin{onlyenv}<+>
      \draw [decorate,decoration={brace,amplitude=10pt,mirror,raise=4pt},yshift=0pt]
      (6.5,2.5) -- (6.5,16) node[draw=black] (sp1) [black,midway,xshift=0.8cm] {};
      \draw node[right=0cm of sp1] { stack pointer };
      \draw[fill=white] (0,2.5) rectangle (6,6.5) node[pos=.5] {\footnotesize current stack frame};
    \end{onlyenv}

    % attacker gets called again + uses old stack pointer.
    \begin{onlyenv}<+>
      \draw[fill=gray!50] (0,2.5) rectangle (6,10.5) node[pos=.5] {\footnotesize caller stack frame};
      \draw[fill=white] (0,10.5) rectangle (6,13.5) node[pos=.5] {\footnotesize callee stack frame};
      \draw (3,12) node[devil,mirrored,minimum size=1cm] {};
      \draw (14,8.25) edge[->,color=red,very thick,out=235,in=0] node[sloped, below] {read/write} (6,8);
      \draw [decorate,decoration={brace,amplitude=10pt,mirror,raise=4pt},yshift=0pt]
        (6.5,10.5) -- (6.5,16) node[draw=black] (sp1) [black,midway,xshift=0.8cm] {};
      \draw node[right=0cm of sp1] { stack pointer };
      % \draw [decorate,decoration={brace,amplitude=10pt,mirror,raise=4pt},yshift=0pt]
      % (6,2.5) -- (6,16) node (callerframe1) [draw=black,midway,xshift=0.8cm] {};
      % \draw node[fill=gray!50,right=0cm of callerframe1] (rp1l) {return pointer data};
    \end{onlyenv}
  \end{tikzpicture}
\end{frame}

\begin{frame}
  \frametitle{Stack and return capabilities: Attack 2}
  \begin{tikzpicture}[scale=.5, every node={scale=.5}]
    % recurrent parts
    \begin{scope}
      \clip (-.1,-.1) rectangle (6.1,15.1);
      \draw[fill=gray!20,draw=none,opacity=.8] (0,0) rectangle (6,15);
      \draw[draw=gray!80] (0,0) -- (0,15);
      \draw[draw=gray!80] (6,0) -- (6,15);
      \draw[fill=gray!50] (0,-.5) rectangle (6,2.5) node[pos=.5,color=black] {\footnotesize higher stack frames...};
      \draw[fill=white] (0,4) rectangle (6,16);
    \end{scope}

    \draw[->] (-2,0) -- node[midway,sloped,above] {stack grows upward} (-2,15);

    % first step: capabilities
    \begin{onlyenv}<+>
      \draw[fill=white] (0,2.5) rectangle (6,4) node[pos=.5] {\footnotesize current stack frame};
      \draw [decorate,decoration={brace,amplitude=10pt,mirror,raise=4pt},yshift=0pt]
        (6.5,2.5) -- (6.5,16) node[draw=black] (sp1) [black,midway,xshift=0.8cm] {};
      \draw node[right=0cm of sp1] { stack pointer };
      % \draw [decorate,decoration={brace,amplitude=10pt,mirror,raise=4pt},yshift=0pt]
      % (6,0) -- (6,16) node (oldframe1) [black,midway,xshift=0.5cm] {};
      % \draw node[fill=gray!50,right=0cm of oldframe1] (rp1l) {return pointer data};
    \end{onlyenv}

    % attacker stores copy of stack pointer high in the stack
    \draw<+->[fill=gray!50] (0,2.5) rectangle (6,4) node[pos=.5] {\footnotesize caller stack frame};
    \begin{onlyenv}<.-.(2)>
      \draw[fill=white] (0,4) rectangle (6,7) node[pos=.5] {\footnotesize callee stack frame};
      \draw (3,5.5) node[devil,mirrored,minimum size=1cm] {};
      \draw [decorate,decoration={brace,amplitude=10pt,mirror,raise=4pt},yshift=0pt]
        (6.5,4) -- (6.5,16) node[draw=black] (sp1) [black,midway,xshift=0.8cm] {};
      \draw node[right=0cm of sp1] { stack pointer };
    \end{onlyenv}
    \draw<+->[fill=white] (0,13) rectangle (6,13.5) node[pos=.5,color=red] {\footnotesize copy of sp};
    \draw<.> (6,13.25) edge[red,thick,->,in=45] (sp1);
    \begin{onlyenv}<+->
      \draw [decorate,decoration={brace,amplitude=10pt,mirror,raise=4pt},yshift=0pt]
        (13.5,4) -- (13.5,16) node[draw=black] (spold) [black,midway,xshift=0.8cm] {};
      \draw (6,13.25) edge[red,thick,->,in=45] (spold);
    \end{onlyenv}

    % attacker calls trusted code again
    \draw<+->[fill=gray!50] (0,4) rectangle (6,7) node[pos=.5] {\footnotesize callee stack frame};
    \begin{onlyenv}<.>
      \draw[fill=white] (0,7) rectangle (6,9) node[pos=.5] {\footnotesize callee (2) stack frame};
      \draw [decorate,decoration={brace,amplitude=10pt,mirror,raise=4pt},yshift=0pt]
        (6.5,7) -- (6.5,16) node[draw=black] (sp1) [black,midway,xshift=0.8cm] {};
      \draw node[right=0cm of sp1] { stack pointer };
    \end{onlyenv}

    % attacker uses old stack pointer
    \draw<+->[fill=gray!50] (0,7) rectangle (6,9) node[pos=.5] {\footnotesize callee (2) stack frame};
    \begin{onlyenv}<.-.(1)>
      \draw[fill=white] (0,9) rectangle (6,12) node[pos=.5] {\footnotesize callee (3) stack frame};
      \draw (3,10.5) node[devil,mirrored,minimum size=1cm] {};
      \draw [decorate,decoration={brace,amplitude=10pt,mirror,raise=4pt},yshift=0pt]
      (6.5,9) -- (6.5,16) node[draw=black] (sp1) [black,midway,xshift=0.8cm] {};
      \draw node[right=0cm of sp1] { stack pointer };
    \end{onlyenv}
    \draw<+> (spold) edge[->,color=red,very thick,out=235,in=0] node[sloped, below] {read/write} (6,8);
  \end{tikzpicture}
\end{frame}

\begin{frame}
  \frametitle{Stack Capabilities Problems}
  \begin{itemize}
  \item Stack capabilities are not a panacea
  \item Stack pointers must not leave stack frame (through memory or stack)
  \item Similar problems for return pointers (omitted)
  \item Solution: local capabilities?
    \begin{itemize}
    \item Make stack and return capabilities local
    \item Cannot leave registers (except on stack) (\sout{Attack 1})
    \item Stack clearing on boundary crossings (\sout{Attack 2})
    \end{itemize}
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Local Stack Capabilities prevent Attack 1}
  \begin{tikzpicture}[scale=.5, every node={scale=.5}]
    % recurrent parts
    \begin{scope}
      \clip (-.1,-.1) rectangle (6.1,15.1);
      \draw[fill=gray!20,draw=none,opacity=.8] (0,0) rectangle (6,15);
      \draw[draw=gray!80] (0,0) -- (0,15);
      \draw[draw=gray!80] (6,0) -- (6,15);
      \draw[fill=gray!50] (0,-.5) rectangle (6,2.5) node[pos=.5,color=black] {\footnotesize higher stack frames...};
    \end{scope}

    \draw[->] (-2,0) -- node[midway,sloped,above] {stack grows upward} (-2,15);

    % attacker...
    \draw<+>[fill=gray!50] (0,2.5) rectangle (6,6.5) node[pos=.5] {\footnotesize caller stack frame};
    \draw<.>[fill=white] (0,6.5) rectangle (6,9.5) node[pos=.5] {\footnotesize callee stack frame};
    \begin{scope}
      \clip (-.1,-.1) rectangle (6.1,15);
      \draw<.>[fill=white] (0,9.5) rectangle (6,15.1);
    \end{scope}
    \draw<.> (3,8) node[devil,mirrored,minimum size=1cm] {};
    \draw<.> [decorate,decoration={brace,amplitude=10pt,mirror,raise=4pt},yshift=0pt]
      (6.5,6.5) -- (6.5,16) node[draw=black] (sp1) [black,midway,xshift=0.8cm] {};
    \draw<.> node[right=0cm of sp1] { stack pointer };

    % stores stack pointer in heap
    \begin{scope}
      \clip (13.9,6) rectangle (18.1,13);
      \draw<.>[fill=gray!20] (14,5.9) rectangle (18,13.1);
    \end{scope}
    \draw<.> (16,14) node {heap memory};
    \draw<.>[fill=gray!50] (14,8) rectangle node[color=red] {\scriptsize copy of old sp} (18,8.5);
    \draw<.> (14,8.25) edge[->,red,thick,bend left] (sp1);

    \draw<.>[green,thick] (13,7.5) -- (19,9);
    \draw<.>[green,very thick] (19,7.5) -- (13,9);
    \draw node[green,cloud,cloud puffs=10.8,cloud puff arc=110, aspect=3, draw] () at (14,4) {Stack pointer is local!};

  \end{tikzpicture}
\end{frame}

\begin{frame}
  \frametitle{Stack Clearing prevent Attack 2}
  TODO
\end{frame}
\end{document}

>>>>

\usepackage{ifthen}
\usepackage{pgf}
\usepackage{ulem}
\usepackage{tikz}
\usetikzlibrary{backgrounds}
\usetikzlibrary{calc}
\usetikzlibrary{positioning}
\usetikzlibrary{patterns}
\usetikzlibrary{decorations.pathreplacing}
\usetikzlibrary{arrows,shapes}
\usepackage{tikzpeople}
\usepackage{amssymb}

% Active adversary
\newcommand{\actadv}[4][]{
  \ifthenelse{\equal{#1}{}}{
    \draw[fill=white] #2 rectangle #3 node[pos=.5] {};
    \draw ($#2 + (3,1.5)$) node[devil,mirrored,minimum size=1cm] {};
    \draw[fill=white, fill opacity=0.5] #2 rectangle #3 node[pos=.5] {};
    \draw #2 rectangle #3 node[pos=.5] {\footnotesize #4};
  }{
    \onlyenv<#1>
    \draw[fill=white] #2 rectangle #3 node[pos=.5] {};
    \draw ($#2 + (3,1.5)$) node[devil,mirrored,minimum size=1cm] {};
    \draw[fill=white, fill opacity=0.5] #2 rectangle #3 node[pos=.5] {};
    \draw #2 rectangle #3 node[pos=.5] {\footnotesize #4};
    \endonlyenv
  }
}

% Inactive adversary
\newcommand{\inactadv}[4][]{
  \ifthenelse{\equal{#1}{}}{
    \draw[fill=white] #2 rectangle #3 node[pos=.5] {};
    \draw ($#2 + (3,1.5)$) node[devil,mirrored,minimum size=1cm] {};
    \draw[fill=white, fill opacity=0.4] #2 rectangle #3 node[pos=.5] {};
    \draw[fill=gray, fill opacity=0.5] #2 rectangle #3 node[pos=.5] {};
    \draw #2 rectangle #3 node[pos=.5] {\footnotesize #4};
  }{
    \onlyenv<#1>
    \draw[fill=white] #2 rectangle #3 node[pos=.5] {};
    \draw ($#2 + (3,1.5)$) node[devil,mirrored,minimum size=1cm] {};
    \draw[fill=white, fill opacity=0.4] #2 rectangle #3 node[pos=.5] {};
    \draw[fill=gray, fill opacity=0.5] #2 rectangle #3 node[pos=.5] {};
    \draw #2 rectangle #3 node[pos=.5] {\footnotesize #4};
    \endonlyenv
  }
}

% Active stack frame
\newcommand{\actsf}[4][]{
  \ifthenelse{\equal{#1}{}}{
    \draw[fill=white] #2 rectangle #3 node[pos=.5] {\footnotesize #4};
  }{
   \draw<#1>[fill=white] #2 rectangle #3 node[pos=.5] {#4};
 }
}

% Inactive stack frame
\newcommand{\inactsf}[4][]{
  \ifthenelse{\equal{#1}{}}{
    \draw[fill=white] #2 rectangle #3 node[pos=.5] {};
    \draw[fill=gray, fill opacity=0.5] #2 rectangle #3 node[pos=.5] {};
    \draw #2 rectangle #3 node[pos=.5] {\footnotesize #4};
  }{
    \draw<#1>[fill=white] #2 rectangle #3 node[pos=.5] {};
    \draw<#1>[fill=gray, fill opacity=0.5] #2 rectangle #3 node[pos=.5] {};
    \draw<#1> #2 rectangle #3 node[pos=.5] {\footnotesize #4};
  }
}

\newcommand{\capbrace}[3][sp1]{
  \draw [decorate,decoration={brace,amplitude=10pt,mirror,raise=4pt},yshift=0pt]
  #2 -- #3 node[draw=black] (#1) [black,midway,xshift=0.8cm] {};

}

\newcommand{\stdstackstart}{
  \scope
    \clip (-.1,-.1) rectangle (6.1,15.1);
    \fill[fill=white] (0,0) rectangle (6,15);
    \draw (0,0) -- (0,15);
    \draw (6,0) -- (6,15);
    \draw[fill=gray!50] (0,-.5) rectangle (6,2.5) node[pos=.5,color=black] {\footnotesize higher stack frames...};
  \endscope
  \draw[->] (-2,0) -- node[midway,sloped,above] {stack grows upward} (-2,15);
}

@@
